---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

set.seed(87346)
```

# nnadic

<!-- badges: start -->
<!-- badges: end -->

The goal of nnadic (**N**eural **N**etwork for **A**symptotic
**D**ependence/**I**ndependence **C**lassification) is to classify
bivariate data sets as either asymptotically dependent or independent
using a trained convolutional neural network. The tool is set up
to automatically:  
- transform the marginal distribution  
- take the top 5% of the data (using the $l_\infty$-norm)  
- resample or subsample as necessary to ensure the data is of the
correct dimension  
- output the predicted result from the neural network

There are many customization options which allow the user to take
advantage of advanced functionality.
    
## Installation

You can install the development version of nnadic from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("twixson/nnadic")
```

Troubleshooting:

- The `nnadic` package requires the `nnadicTestData` package which has
  290Mb of data in it. Some users have found that they need to run
  `options(timeout = 400)` prior to installation.  
- The `nnadic` package requires a working `keras` installation. Some
  users have found that installing `keras` first is helpful. This
  includes ensuring that the python installation needed for the
  creation of the virtual environment is suitable. 

Suggested install code:

``` r
install.packages("remotes", force = TRUE)
remotes::install_github("rstudio/tensorflow", force = TRUE)
remotes::install_github("rstudio/keras", force = TRUE)
reticulate::install_miniconda()
tensorflow::install_tensorflow()
reticulate::install_python(version = '3.9')
keras::install_keras()
options(timeout = 400)
install_github("twixson/nnadicTestData", force = TRUE)
install_github("twixson/nnadic")
```


## Example

This is a basic example which shows you the common workflow:

```{r example}
library(nnadic)
library(nnadicTestData)
library(evd)      # for generating ADep datasets (logistic)
library(mvtnorm)  # for generating AInd datasets (gaussian)

test_data_four <- make_symmetric(test_data_four)

results <- nnadic(test_data_four, one_test = FALSE)
mean(results$preds == test_response_four)
```


```{r}
set.seed(2938)
# Lets generate a new Adep dataset
data <- rbvevd(8734, dep = 0.5, model = "log")
# and test it
data_ready <- get_nnadic_input(data)
results <- nnadic(data_ready)
# should be 0
```

```{r}
# Lets generate a new Adep dataset
data <- rmvnorm(19886, c(0,0), matrix(c(1, 0.5, 0.5, 1), nrow = 2))
# and test it
data_ready <- get_nnadic_input(data, subsample = TRUE)
results <- nnadic(data_ready, make_hist = FALSE)
# should be 1
```

